Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-01-19 writeup/pnas-word-order-si.tex
reproduced per-dependency results

@@

We employ two common variance reduction methods to improve the estimator~(\ref{eq:estimator}), while keeping it unbiased.
For dependency length and predictability, note that the loss incurred for a specific word only depends on ordering decisions made up to that word (and its head, in the case of dependency length). We represent the process of linearizing a tree as a dynamic stochastic computation graph, and use these independence properties to apply the method described in \citet{schulman2015gradient} to obtain a version of~(\ref{eq:estimator}) with lower variance.
Second, we use a word-dependent moving average of recent per-word losses as control variate \cite{williams1992simple}.
These two methods reduce the variance of the estimator and thereby increase the speed of optimization and reduce training time, without biasing the results.
For numerical stability, we represent $a_\tau \in [0,1]$ via its logit $\in \mathbb{R}$.
Furthermore, to encourage exploration of the parameter space, we add an entropy penalty \citep{xu2015show} for each Direction Parameter $a_\tau$, which penalizes $a_\tau$ values near $0$ or $1$. The weight of the entropy penalty was chosen together with the other hyperparameters.
%, adding the negative entropy of the $Bernoulli(a_\tau)$ to the loss $R$. %, weighted with $\alpha$.

@@

We update word order grammar parameters $\theta$ using SGD with momentum.
For the language model parameters $\phi$, we use plain SGD. 
For the parser parameters $\psi$, we use Adam \citep{kingma2014adam}, following \citet{dozat2017stanford}.
The learning rates and momentum weight of SGD, and the parameters of Adam were determined together with the other hyperparameters.

For each language and objective function, we apply early stopping using a Monte-Carlo estimate of the respective objective functions on the UD development set. We compute an estimate of the objective on the development set in intervals of 50,000 training steps, and stop optimization once the value of the objective deteriorates compared to the last previous estimate. This technique, well known in the machine learning literature, helps prevent overfitting to the specific corpora used.

>>>>

A different commonly considered information-theoretic objective function for word order is Uniform Information Density~\citep{jaeger2010redundancy}, formalized as minimizing a superlinear function of per-word surprisal~\citep{levy:2018cogsci}.
Note that optimizing grammars for this objective faces additional challenges, as the optimization problem cannot be rewritten as joint optimization in the form (\ref{eq:maximize-pred}).
We thus do not evaluate this objective function here.

@@

We employ two common variance reduction methods to improve the estimator~(\ref{eq:estimator}), while keeping it unbiased.
For dependency length and predictability, note that the loss incurred for a specific word only depends on ordering decisions made up to that word (and its head, in the case of dependency length). We represent the process of linearizing a tree as a dynamic stochastic computation graph, and use these independence properties to apply the method described in \citet{schulman2015gradient} to obtain a version of~(\ref{eq:estimator}) with lower variance.
Second, we use a word-dependent moving average of recent per-word losses as control variate \cite{williams1992simple}.
These two methods reduce the variance of the estimator and thereby increase the speed of optimization and reduce training time, without biasing the results.
For numerical stability, we represent $a_\tau \in [0,1]$ via its logit $\in \mathbb{R}$.
Furthermore, to encourage exploration of the parameter space, we add an entropy regularization term \citep{xu2015show} for each Direction Parameter $a_\tau$, which penalizes $a_\tau$ values near $0$ or $1$. The weight of the entropy regularization was chosen together with the other hyperparameters.
%, adding the negative entropy of the $Bernoulli(a_\tau)$ to the loss $R$. %, weighted with $\alpha$.

@@

We update word order grammar parameters $\theta$ using SGD with momentum.
For the language model parameters $\phi$, we use plain SGD. 
For the parser parameters $\psi$, we use Adam \citep{kingma2014adam}, following \citet{dozat2017stanford}.
The learning rates and other optimization hyperparameters were determined together with the other hyperparameters.

For each language and objective function, we apply Early Stopping~\citep{prechelt1998early} as follows. We compute a Monte-Carlo estimate of the objective on the UD development set in intervals of 50,000 training steps, and stop optimization once the value of the objective deteriorates compared to the last previous estimate. This technique, well known in the machine learning literature, helps prevent overfitting to the specific corpora used.
